\Huge{Machine Learning} \\ \LARGE Nature's Quest: Deep Learning Exploration of Global Plant Traits from Images and Geodata

### 2024-03-29 paper/paper.tex
result plots

\smallskip
For our method, we wanted to compare the accuracy of a CNN trained on images alone and using a pre-trained backbone (e.g. ResNet, EfficientNet, etc.) to a CNN trained on images and geodata. We hypothesize that the CNN trained on images and geodata will outperform the CNN trained on images alone. Although the geodata could potentially lead the model to learning features not necessarily helpful for accurate prediction of plant traits due to a multitude of reasons.

@@

Integral to any machine learning model is the data used to train, validate, and ultimately test the model. The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxiliary task standard deviations, and the training images of the plants.

@@

\smallskip 
One of the first things we decided to test was the learning rate schedule. We tested out 3 different schedules: step-based (baseline), exponential, and cosine annealing. The results demonstrated in \ref{fig:lr_schedules} show that the cosine annealing and exponential schedule had a slower convergence than the step-based schedule. This is likely due to the fact that the step-based schedule is more aggressive in its learning rate decay.

\begin{figure}[!h]
    \centering
    \includegraphics[width=1\textwidth]{assets/lr_schedules.png}
    \caption{Comparison of different learning rate schedules}
    \label{fig:lr_schedules}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth]{assets/train_vs_notrain.png}
    \caption{Comparison of of models with backbone training on and off}
    \label{fig:backbone_training}
\end{figure}

@@

\begin{figure}[!h]
    \centering 
    \includegraphics[width=0.8\textwidth]{assets/img_geo_model.png}
    \caption{Model architecture of the CNN trained on images and geodata}
    \label{fig:geodata_model}
\end{figure}

@@

% discuss the results
The prediction of plant traits from images and geodata is a complex task that requires a CNN to learn and extract featues from both images and geodata. The results of our work show that a CNN trained on images and geodata outperforms a CNN trained on images alone. Though neither of the models elaborated here is necessarily a good predictior, one of the main reasons for this is likely the complexity and nuance of the task. Though in addition there are various ways our method could be improved upon.

@@

How adaptable pretrained models are to new tasks is also is also something that could be further explored. Kunze et al. \cite{kunze2017transfer} answered this question by looking at the weight difference distribution per layer early and late into training. What they found was that for models trained on a similar task to the target domain only had a minor change in the weight distribution. This reasoning could possibly be applied to more accurately find the optimal backbone. Additionally the use of a pretrained model could be further explored by looking at the effect of the number of layers frozen during training.

\section{Conclusion}

% give some future direction for things to go into

>>>>

\smallskip
For our method, we wanted to compare the accuracy of a CNN trained on images alone and using a pre-trained backbone (e.g. ResNet, EfficientNet, etc.) to a CNN trained on images and geodata. We hypothesize that the CNN trained on images and geodata will outperform the CNN trained on images alone. With our results we hope to provide a better understanding of the key factors that are needed to predict plant traits.

@@

Integral to any machine learning model is the data used to train, validate, and ultimately test the model. The data used consisted of 3 main components: ancillary geodata from various sources, the main task trait means and auxiliary task standard deviations, and the training ima ges of the plants.

@@

\smallskip 
One of the first things we decided to test was the learning rate schedule. We tested out 3 different schedules: step-based (baseline), exponential, and cosine annealing. The results demonstrated in \ref{fig:lr_schedules} show that the cosine annealing and exponential schedule had a slower convergence than the step-based schedule and a more erratic behavior.

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.4\textwidth]{assets/lr_schedule_diffs.png}
    \caption{Comparison of different learning rate schedules}
    \label{fig:lr_schedules}
\end{figure}

@@

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth]{assets/train_vs_notrain.png}
    \caption{Comparison of models with backbone training on (right) and off (left)}
    \label{fig:backbone_training}
\end{figure}

@@

\begin{figure}[!h]
    \centering 
    \includegraphics[width=0.8\textwidth]{assets/img_geo_model.png}
    \caption{(right) Model architecture of the CNN trained on images and geodata; (left) Model architecture of the CNN trained on images alone}
    \label{fig:geodata_model}
\end{figure}

@@

% discuss the results
The prediction of plant traits from images and geodata is a complex task that requires a CNN to learn and extract featues from both images and geodata. The results of our work show that a CNN trained on just image outpreforms one trained on images and geodata. Though neither of the models elaborated here is necessarily a good predictior, one of the main reasons for this is likely the complexity and nuance of the task. We now propose various directions for future work to improve the prediction capabilities of the models.

@@

How adaptable pretrained models are to new tasks is also is also something that could be further explored. Kunze et al. \cite{kunze2017transfer} answered this question by looking at the weight difference distribution per layer early and late into training. What they found was that models trained on a similar task to the target domain only had a minor change in the weight distribution. This reasoning could possibly be applied to more accurately find the optimal backbone. Additionally the use of a pretrained model could be further explored by looking at the effect of the number of layers frozen during training.

\section{Conclusion}
To conclude we noticed that first and foremost the task to predict pant traits is a hard one. It requires a careful consideration during every step of the process from understanding the data to how a neural network architecture will learn from it. From our results we observed that a model trained on images and ancillary geodata converges faster but does not necessarily have a higher $R^2$ than a model trained on images alone. In the final tests run these results were also observed with our image model having a higher $R^2$ than the combined model. Though both models still have a negative $R^2$ suggesting some core limitations in the models. Future work could look at the effect of different hyperparameters, model architectures, and data modifications to improve the prediction capabilities of the models.
